---
output: html_document
editor_options: 
  chunk_output_type: console
---
# Geography

CCDPH's jurisdiction is within the largely urban Bellwood health region, one of [seven health regions within the state of Illinois](https://dph.illinois.gov/contact-us/regional-health-departments.html). The Bellwood health region, which corresponds with the Cook County boundary, is the smallest of the state's health regions in terms of area, although it has the most residents, comprising over 40 percent of the total population of Illinois.

CCDPH is one of six local health departments within the Bellwood health region. The other five local health departments are the [Chicago Department of Public Health](https://www.chicago.gov/city/en/depts/cdph.html), the [City of Evanston Department of Health and Human Services](https://www.cityofevanston.org/government/departments/health-human-services), the [Village of Oak Park Department of Public Health](https://www.oak-park.us/village-services/department-public-health), the [Stickney Township Public Health District](http://stickneypublichealthdistrict.org/) and the [Village of Skokie Health Department](https://www.skokie.org/158/Health). The [latest contact information for local health departments in Illinois](https://idph.illinois.gov/IDPHPrograms/v_LHDDirectory/Show-V-LHDDirectory-Public.aspx?RegionServed=7) can be found on the Illinois Department of Public Health website.

# CCDPH-relevant boundary shapefiles
The remainder of this document describes the geographic boundary layers that are critical to CCDPH's epidemiology work. The described layers are available in the *inter-spatial* database on the CCDPH SQL Server. All spatial data in the database are stored with the epsg 3435 coordinate reference system (CRS).

Below is code that was used to process CCDPH-relevant cartographic boundary shapefiles made available in the CCDPH *inter-spatial* database. The geographic data represent both political and analytical boundaries derived from the US Census Bureau via its [TIGER/Line](https://www.census.gov/geographies/mapping-files/time-series/geo/tiger-line-file.html) program. By default, census-derived shapes are generalized to 1:500K. Set "cb=FALSE" to acquire more detailed TIGER/Line files.

# R functions for reading/writing spatial data from/to database
The below functions can be used to efficiently read/write spatial data from/to the CCDPH SQL Server database using native SQL Server geometry types. The function is also available on the [ccdph-functions GitHub repository](https://github.com/Cook-County-Department-of-Public-Health/ccdph-functions).

# Required packages
Activate the following packages if it is necessary to read, modify or re-create any of the boundary files. Also remember that the SQL Server name is to be encrypted in order to improve privacy of databases.

```{r}
#| label: required packages

library(sf)
library(DBI)
library(odbc)
library(dplyr)
library(tidyverse)
library(spData)
library(keyring)
library(tigris)
library(units)
```

###Read spatial data from inter-spatial database

```{r}
# read in latest spatial data functions
 source("https://raw.githubusercontent.com/Cook-County-Department-of-Public-Health/ccdph-functions/master/spatial-database-functions.R")

# example
test <- fx_read_spatial_layer_fr_database(sf_layer_name = "counties_il_new",
                                  db_table_name = "counties_illinois",
                                  crs_id = 4326)
```

###Write spatial data from inter-spatial database

```{r}
# write latest spatial data functions
 source("https://raw.githubusercontent.com/Cook-County-Department-of-Public-Health/ccdph-functions/master/spatial-database-functions.R")

test <- fx_read_spatial_layer_fr_database(sf_layer_name = "counties_il_new",
                                  db_table_name = "counties_illinois",
                                  crs_id = 4326)

```

###Create, format and write boundary data to inter-spatial database
## Illinois county boundaries file
There are 102 counties in Illinois. The code below imports, formats and writes county boundaries (including Cook County [geoid = "17031"]) to the inter-spatial database.

```{r}
#| label: Illinois counties

# download latest data from US Census TIGER/Line files
counties_il <- counties(state="17", cb=TRUE)

# reformat data, standardize column names
counties_il <- counties_il %>%
  st_transform(3435) %>%
  st_as_sf() %>%
  select(geoid_county = GEOID,
         name_county = NAME,
         area_land = ALAND,
         area_water = AWATER) %>%
  mutate(area_land = set_units(area_land,ft^2),
         area_water = set_units(area_water,ft^2),
         area_land_sqmi = set_units(area_land, mi^2),
         area_water_sqmi = set_units(area_water,mi^2),
         area_sqmi = area_land_sqmi + area_water_sqmi) %>%
  select(-c(area_land,area_water))

# write data to database
counties_il_db <- fx_spatial_layer_to_database(
  sf_layer_name = counties_il,
  schema_name = "ref",
  db_table_name = "counties_illinois",
  crs_id = 3435)

```

## ZCTAs in Cook County
Zip code tabulation areas (ZCTAs) in Cook County.
```{r}
#| label: ZCTas in Cook County

# download latest data from US Census TIGER/Line files 
# (2020 data not stratified by state)
zctas_us <- zctas(cb=TRUE, year=2020) %>%
  st_drop_geometry() %>%
  select(geoid_zcta = GEOID20,
         area_land = ALAND20,
         area_water = AWATER20)

# download latest data from ccdph-shapefiles GitHub 
zctas_cook_county <- st_read("C:/Users/christopher.smith/OneDrive - Cook County Health/git_repos/justenvirons/ccdph-shapefiles/epsg-3435-illinois-stateplane-east/cook_county_zctas_epsg3435.shp") %>%
  select(geoid_zcta = zcta,
         primary_city = prmry_c,
         partial_ccdph = prtl_pp,
         percent_ccdph = prcnt_d,
         notes) %>%
  drop_na(geoid_zcta) %>%
  left_join(zctas_us, by="geoid_zcta") %>%
  st_as_sf() %>%
  mutate(area_land = set_units(area_land,ft^2),
         area_water = set_units(area_water,ft^2),
         area_land_sqmi = set_units(area_land, mi^2),
         area_water_sqmi = set_units(area_water,mi^2),
         area_sqmi = area_land_sqmi + area_water_sqmi) %>%
  select(-c(area_land, area_water)) %>%
  st_transform(crs = 3435) %>%
  st_make_valid()

# write simple feature layer to database
zctas_cook_county_db <- fx_spatial_layer_to_database(
  sf_layer_name = zctas_cook_county,
  schema_name = "ref",
  db_table_name = "zctas_cook_county",
  crs_id = 3435)

munis_cook_county_db <- fx_write_spatial_layer_to_database(
  sf_layer_name = zctas_cook_county,
  db_table_name = "zctas_cook_county")

```

## ZCTAs in Cook County (clipped)
```{r}
# clip zctas_cook_county data in ArcGIS Pro 

zctas_cook_county_clipped <- st_read("C:/Users/christopher.smith/OneDrive - Cook County Health/git_repos/justenvirons/ccdph-jurisdictions/layers/zctas_cook_county_clipped.shp") 

zctas_cook_county_clipped_db <- fx_write_spatial_layer_to_database(
  sf_layer_name = zctas_cook_county_clipped,
  db_table_name = "zctas_cook_county_clipped")
```


## Municipalities in Cook County

```{r}
#| label: municipalities in Cook County

# download latest data from US Census TIGER/Line files
places_il <- places(state = "17", cb=TRUE) %>% 
  st_as_sf() %>%
  st_transform(crs = 3435)

# download latest data from ccdph-shapefiles GitHub 
munis_cook_county <- st_read("C:/Users/christopher.smith/OneDrive - Cook County Health/git_repos/justenvirons/ccdph-shapefiles/epsg-3435-illinois-stateplane-east/ccdph-cook-county-munis.shp") %>%
  st_drop_geometry() %>%
  drop_na(muni_name) %>%
  left_join(places_il %>% 
              select(NAME,geoid_place = GEOID,area_land = ALAND,area_water= AWATER), 
            by=c("muni_name"="NAME")) %>%
  st_as_sf() %>%
  mutate(area_land = set_units(area_land,ft^2),
         area_water = set_units(area_water,ft^2),
         area_land_sqmi = set_units(area_land, mi^2),
         area_water_sqmi = set_units(area_water,mi^2),
         area_sqmi = area_land_sqmi + area_water_sqmi,
         location = case_when(district=="OOJ" & muni_name!="Chicago" ~ "SCC",
                              district=="OOJ" & muni_name=="Chicago" ~ "Chicago",
                              district != "OOJ" ~ "CCDPH")) %>%
  select(name_place = muni_name,
         geoid_place,
         location,
         district,
         partial_ccdph = partial,
         exclude,
         area_land_sqmi,
         area_water_sqmi,
         area_sqmi) %>%
  arrange(name_place) %>%
  st_make_valid()

# write simple feature layer to database
munis_cook_county_db <- fx_write_spatial_layer_to_database(
  sf_layer_name = munis_cook_county,
  db_table_name = "munis_cook_county")

```

## Municipalities in Cook County (clipped)
```{r}
# clip munis_cook_county data in ArcGIS Pro 

munis_cook_county_clipped <- st_read("C:/Users/christopher.smith/OneDrive - Cook County Health/git_repos/justenvirons/ccdph-jurisdictions/layers/munis_cook_county_clipped.shp") 

munis_cook_county_clipped_db <- fx_write_spatial_layer_to_database(
  sf_layer_name = munis_cook_county_clipped,
  db_table_name = "munis_cook_county_clipped")
```

## CCDPH jurisdiction
```{r}
#| label: CCDPH jurisdiction boundary

# download latest data from US Census TIGER/Line files
IL_Places_geom <- places(c("IL"), cb=TRUE, class="sf", year=2020)
IL_Counties_geom <- counties(c("IL"), cb=TRUE, class="sf", year=2020)
IL_MCDs_geom <- county_subdivisions(c("IL"), cb=TRUE, class="sf", year=2020)
IL_Places_geom <- st_transform(IL_Places_geom, crs = 3435, year=2020)
IL_Counties_geom <- st_transform(IL_Counties_geom, crs = 3435, year=2020)
IL_MCDs_geom <- st_transform(IL_MCDs_geom, crs = 3435, year=2020)

# Select out of jurisdiction (OOJ) places
ooj_places <- IL_Places_geom %>% 
  filter(#NAME=="Chicago" |
           # NAME=="Bedford Park" |
           # NAME=="Stickney" |
           # NAME=="Burbank" |
           # NAME=="Forest View" |
           NAME=="Evanston" | 
           NAME=="Oak Park" | 
           NAME=="Skokie"
           ) %>%
  select(GEOID_place = GEOID,
         NAME_place = NAME)

# Select Minor Civil Divisions (MCD) Stickney Township, Cook County
ooj_mcds <- IL_MCDs_geom %>% 
  filter(COUNTYFP=="031" &
           (NAME=="Stickney" |
           NAME=="Chicago")) %>%
  select(NAME_mcd = NAME,
         GEOID_mcd = GEOID)

# Create mask from union of OOJ places and Minor Civil Divisions --------
ooj_mask <-  ooj_places %>% 
  st_union(ooj_mcds, by_feature = FALSE) %>% 
  st_union(by_feature = FALSE) %>% st_as_sf()

# Select Cook County boundary
cc_boundary <- IL_Counties_geom %>% 
  filter(STATEFP=="17" & COUNTYFP=="031") %>%
  select(GEOID_county = GEOID,
         NAME_county = NAME)

# Use mask to clip from Cook County and produce final CCDPH boundary
ccdph_boundary <- st_difference(cc_boundary,
                                        ooj_mask) %>%
  mutate(NAME_ccdph="CCDPH boundary") %>%
  select(NAME_ccdph) %>%
  mutate(area_sqmi = st_area(geometry),
         area_sqmi = set_units(area_sqmi, ft^2),
         area_sqmi = set_units(area_sqmi, mi^2))


# write simple feature layer to database
ccdph_boundary_db <- fx_spatial_layer_to_database(
  sf_layer_name = ccdph_boundary,
  schema_name = "ref",
  db_table_name = "ccdph_boundary",
  crs_id = 3435)

```

## CCDPH districts 
### District boundaries for municipality-level mapping

```{r}
# download latest data from ccdph-shapefiles GitHub 
ccdph_district_boundaries_muni <- st_read("C:/Users/christopher.smith/OneDrive - Cook County Health/git_repos/justenvirons/ccdph-shapefiles/epsg-3435-illinois-stateplane-east/ccdph_districts_boundaries_muni_epsg3435.shp") %>%
  mutate(district = if_else(district=="Outside", "OOJ", district),
         area_sqmi = st_area(geometry),
         area_sqmi = set_units(area_sqmi, ft^2),
         area_sqmi = set_units(area_sqmi, mi^2))

# write simple feature layer to database
ccdph_district_boundaries_muni_db <- fx_spatial_layer_to_database(
  sf_layer_name = ccdph_district_boundaries_muni,
  schema_name = "ref",
  db_table_name = "ccdph_district_boundaries_muni",
  crs_id = 3435)

```

### District boundaries for census tract-level mapping

```{r}
# download latest data from ccdph-shapefiles GitHub 
ccdph_district_boundaries_tract <- st_read("C:/Users/christopher.smith/OneDrive - Cook County Health/git_repos/justenvirons/ccdph-shapefiles/epsg-3435-illinois-stateplane-east/ccdph_districts_boundaries_tracts_epsg3435.shp") %>%
  select(-sqmi) %>%
  rename(num_tracts_2020 = tracts,
         name_district = district) %>%
  mutate(name_district = str_to_title(name_district),
         name_district = if_else(name_district=="Outside", "OOJ", name_district),
         area_sqmi = st_area(geometry),
         area_sqmi = set_units(area_sqmi, ft^2),
         area_sqmi = set_units(area_sqmi, mi^2))

# write simple feature layer to database
ccdph_district_boundaries_tract_db <- fx_spatial_layer_to_database(
  sf_layer_name = ccdph_district_boundaries_tract,
  schema_name = "ref",
  db_table_name = "ccdph_district_boundaries_tract",
  crs_id = 3435)

```

## Census tracts
### 2010 census tracts in Cook County

```{r}
# download latest data from US Census TIGER/Line files
tracts_2010_cook_county_census <- tracts(state="17", county="031", cb=TRUE, year=2010) %>%
  st_drop_geometry() %>%
  mutate(geoid_tract_2010 = paste0(STATE,COUNTY,TRACT),
         name_tract_2010 = NAME,
         area_sqmi = set_units(CENSUSAREA, mi^2)) %>%
  select(geoid_tract_2010,
         name_tract_2010,
         area_sqmi)

# download latest data from ccdph-shapefiles GitHub 
tracts_2010_cook_county <- st_read("C:/Users/christopher.smith/OneDrive - Cook County Health/git_repos/justenvirons/ccdph-shapefiles/epsg-3435-illinois-stateplane-east/cc_tracts_2010_epsg3435.shp") %>%
  select(-sqmi) %>%
  rename(geoid_tract_2010 = fips,
         name_district = district) %>%
  mutate(name_district = str_to_title(name_district),
         name_district = if_else(name_district=="Outside", "OOJ", name_district),
         location = str_to_upper(location),
         location = if_else(location=="CHICAGO","Chicago",location)) %>%
  left_join(tracts_2010_cook_county_census, by="geoid_tract_2010")

# write simple feature layer to database
tracts_2010_cook_county_db <- fx_spatial_layer_to_database(
  sf_layer_name = tracts_2010_cook_county,
  schema_name = "ref",
  db_table_name = "tracts_2010_cook_county",
  crs_id = 3435)

```

### 2020 census tracts in Cook County

```{r}
# download latest data from US Census TIGER/Line files
tracts_2020_cook_county_census <- tracts(state="17", county="031", cb=TRUE, year=2020) %>%
  st_drop_geometry() %>%
  select(geoid_tract_2020 = GEOID,
         name_tract_2020 = NAME,
         area_land = ALAND,
         area_water = AWATER) %>%
  mutate(area_land = set_units(area_land,ft^2),
         area_water = set_units(area_water,ft^2),
         area_land_sqmi = set_units(area_land, mi^2),
         area_water_sqmi = set_units(area_water,mi^2),
         area_sqmi = area_land_sqmi + area_water_sqmi) %>%
  select(geoid_tract_2020,
         name_tract_2020,
         area_land_sqmi:area_sqmi)

# download latest data from ccdph-shapefiles GitHub 
tracts_2020_cook_county <- st_read("C:/Users/christopher.smith/OneDrive - Cook County Health/git_repos/justenvirons/ccdph-shapefiles/epsg-3435-illinois-stateplane-east/cc_tracts_2020_epsg3435.shp") %>%
  select(-sqmi) %>%
  rename(geoid_tract_2020 = fips,
         name_district = district) %>%
  mutate(name_district = str_to_title(name_district),
         name_district = if_else(name_district=="Outside", "OOJ", name_district),
         location = str_to_upper(location),
         location = if_else(location=="CHICAGO","Chicago",location)) %>%
  left_join(tracts_2020_cook_county_census, by="geoid_tract_2020")

# write simple feature layer to database
tracts_2020_cook_county_db <- fx_spatial_layer_to_database(
  sf_layer_name = tracts_2020_cook_county,
  schema_name = "ref",
  db_table_name = "tracts_2020_cook_county",
  crs_id = 3435)

```
