# Geography {#geography .unnumbered}
This chapter describes basic characteristics of CCDPH's jurisdiction and other geographies that are used by CCDPH's Epidemiology Unit. The chapter also describes how CCDPH's reference geographic layers are stored and how to access (or import) them via different applications, including R, SAS, and ArcGIS Pro. Note that many of the boundaries discussed in this section change over time due to a variety of reasons including: annexations by cities, re-delineation of administrative and political districts, accuracy updates. A list of the key reference (or base) layers are listed in @fig-spatialtables below.

```{r}
#| eval: true
#| echo: false

library(dplyr)
library(dbplyr)
library(tidyverse)
library(odbc)
library(DBI)
library(DT)
library(data.table)
library(sf)
library(sp)
library(leaflet)
library(leaflet.extras)
library(leaflet.providers)
library(keyring)
library(openxlsx)

source("https://raw.githubusercontent.com/Cook-County-Department-of-Public-Health/ccdph-functions/master/spatial-database-functions.R")
```

```{r}
#| label: fig-spatialtables
#| fig-cap: Geography Tables in inter-spatial SQL Server Database
#| warning: false
#| eval: true

con_inter_spatial <- dbConnect(
  odbc::odbc(),
  Driver   = "SQL Server",
  Server   = key_get("ccdph_sql_server"),
  Database = "inter-spatial"
)

`00_table_dictionary` <- dbReadTable(conn = con_inter_spatial,
                               Id(schema = "ref", 
                                  table = "00_table_dictionary"))

`00_table_dictionary` <- `00_table_dictionary` %>%
  select(name,
         description,
         source,
         vintage)

`00_table_dictionary` %>%
  arrange(name) %>%
    datatable(class = 'cell-border stripe',
            rownames = FALSE,
            options = list(pageLength = 5)) %>%
  formatStyle(columns = c(1:4),
              
              fontSize = '75%')
```

::: {.callout-note collapse="false"}
## Code for creating geographic layers
The Epidemiology and Communicable Disease units develop and share code used to produce and manage geographic layers via the [ccdph-shapefiles](https://github.com/Cook-County-Department-of-Public-Health/ccdph-shapefiles) GitHub repository.
[This Quarto document](https://github.com/Cook-County-Department-of-Public-Health/ccdph-shapefiles/blob/main/shapefile-creation-code/spatial_data_creation_latest.qmd) provides code for developing and writing spatial datasets to the inter-spatial SQL Server database. For more information on how to access geographic data, refer to the [Storage and Access](storage.qmd) section of the handbook.
:::

## CCDPH jurisdiction
Cook County Department of Public Health (CCDPH) was established in 1945 and is the state-certified public health agency for most of suburban Cook County (SCC). CCDPH’s jurisdiction is within the largely urban Bellwood health region, one of [seven health regions within the state of Illinois](https://dph.illinois.gov/contact-us/regional-health-departments.html). CCDPH has an approximate area of 693 square miles and serves 2,288,429 million residents @census2020dhc.

# Geographic boundaries as units of analysis
The Epidemiology Unit reports public health information for many geographic units of analysis or boundaries beyond CCDPH's jurisdiction. Selecting an appropriate geographic unit depends on the scope and focus of the given analysis. For example, a coarse-grained assessment of death rates may report mortality trends for CCDPH's jurisdiction as whole, whereas a fine-grained study of morbidity may examine variations in emergency department visits by municipality or census tract.

What follows are brief descriptions of the most commonly used geographic units or boundaries employed by Epidemiology Unit staff beyond the CCDPH's jurisdiction. For the most part, boundaries are arranged in descending order by size (i.e., area).

## Interactive map of CCDPH-related geographies

@fig-geographies provides an interactive display of the most commonly used geographic units or boundaries used by Epidemiology Unit staff.

```{r}
#| label: fig-geographies
#| fig-cap: CCDPH-Related Geographies
#| eval: true
#| echo: true

# import GIS layers from inter-spatial
ccdph_boundary <- fx_read_spatial_layer_fr_database(db_table_name = "ccdph_boundary", crs_id = 4326)
scc_boundary <- fx_read_spatial_layer_fr_database(db_table_name = "scc_boundary", crs_id = 4326)
counties_illinois <- fx_read_spatial_layer_fr_database(db_table_name = "counties_illinois", crs_id = 4326) %>% dplyr::filter(geoid_county=="17031")
munis_cook_county_clipped <- fx_read_spatial_layer_fr_database(db_table_name = "munis_cook_county_clipped", crs_id = 4326)
ccdph_districts <- fx_read_spatial_layer_fr_database(db_table_name = "ccdph_district_boundaries_muni", crs_id = 4326)

leaflet() %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addFullscreenControl() %>%
  addPolygons(
    group = "Cook County",
    data = counties_illinois,
    fillColor = "orange",
    weight = 2,
    opacity = 0.5,
    color = "white",
    dashArray = "3",
    fillOpacity = 0.7
  ) %>%
  addPolygons(
    group = "CCDPH",
    data = ccdph_boundary,
    weight = 2,
    opacity = 0.5,
    color = "white",
    dashArray = "3",
    fillColor = "darkblue",
    fillOpacity = 0.7
  ) %>%
  addPolygons(
    group = "SCC",
    data = scc_boundary,
    weight = 2,
    opacity = 0.5,
    color = "white",
    dashArray = "3",
    fillColor = "#5097A4",
    fillOpacity = 0.7
  ) %>%
  addPolygons(
    group = "Munis",
    data = munis_cook_county_clipped,
    weight = 2,
    opacity = 0.5,
    color = "white",
    dashArray = "1",
    fillColor = "lightblue",
    fillOpacity = 0.7
  ) %>%
  addPolygons(
    group = "Districts",
    data = ccdph_districts %>% dplyr::filter(name_district !=
                                               "OOJ"),
    weight = 2,
    opacity = 0.5,
    color = "black",
    dashArray = "1",
    fillColor = NULL,
    fillOpacity = 0
  ) %>%
  addLayersControl(
    overlayGroups = c("Cook County", "CCDPH", "SCC", "Munis", "Districts")
    #,
    #options = layersControlOptions(collapsed = FALSE)
  ) %>%
  hideGroup(c("CCDPH", "SCC", "Munis", "Districts"))

```

## Cook County (aka Bellwood health region)
Cook County is one of Illinois' 102 counties and is the largest county in Illinois in terms of population, comprising over 40 percent of the total population of Illinois. The Bellwood health region, which shares the same boundary as Cook County, is the smallest of the state’s health regions in terms of area  (1,635 square miles), although it has the most residents (5.2 million in 2020) and comprises over 40 percent of the total population of Illinois.

CCDPH is one of six local health departments (LHDs) within Cook County. The other five are the [Chicago Department of Public Health](https://www.chicago.gov/city/en/depts/cdph.html), the [City of Evanston Department of Health and Human Services](https://www.cityofevanston.org/government/departments/health-human-services), the [Village of Oak Park Department of Public Health](https://www.oak-park.us/village-services/department-public-health), the [Stickney Township Public Health District](http://stickneypublichealthdistrict.org/) and the [Village of Skokie Health Department](https://www.skokie.org/158/Health). The [latest contact information for local health departments in Illinois](https://idph.illinois.gov/IDPHPrograms/v_LHDDirectory/Show-V-LHDDirectory-Public.aspx?RegionServed=7) can be found on the Illinois Department of Public Health website.

## Suburban Cook County (SCC)
Suburban Cook County or SCC refers to the entire area of Cook County excluding the city of Chicago. It has an area of approximately 727 square miles. Suburban Cook County includes CCDPH's jurisdiction as well as the areas of the other four local health departments. According to the 2020 decennial census, SCC has a population of 2,529,711 @census2020dhc.

## CCDPH health districts
CCDPH's jurisdiction is also divided into four public health districts: north; west; southwest; and south. These four CCDPH public health districts are used by CCDPH largely to evaluate social and demographic patterns across its jurisdiction, as well as to enumerate and report the incidence and prevalence of disease and other health indicators.

```{r}
#| label: fig-ccdphdistricts
#| fig-cap: CCDPH health districts by area and population
#| warning: false
#| eval: true

con_inter_census <- dbConnect(
  odbc::odbc(),
  Driver   = "SQL Server",
  Server   = key_get("ccdph_sql_server"),
  Database = "inter-census"
)

pop_districts_cook_county <- dbReadTable(conn = con_inter_census,
                               Id(schema = "ref", 
                                  table = "decennial-2020-total-by-ccdph-scc-addition"))

ccdph_districts <- dbReadTable(conn = con_inter_spatial,
                               Id(schema = "ref", 
                                  table = "ccdph_district_boundaries_tract"))

ccdph_districts_frm <- ccdph_districts %>%
    select(-c(geom,num_tracts_2020)) %>%
  left_join(pop_districts_cook_county %>% select(name_district = category,population = pop_ccdph), by="name_district") %>%
  dplyr::filter(name_district != "OOJ") %>%
  rename(name = name_district,
         sqmi = area_sqmi)

ccdph_districts_frm %>%
  arrange(name) %>%
    datatable(class = 'cell-border stripe',
            rownames = FALSE,
            options = list(pageLength = 5)) %>%
  formatRound('sqmi',1) %>%
  formatRound(c(3),0, interval = 3, mark =  ",") %>%
  formatStyle(columns = c(1:3),
              fontSize = '75%')

```

## SCC municipalities 
CCDPH’s jurisdiction includes 120 municipalities, 114 in their entirety and portions of six others. The unincorporated area population (i.e., the population residing outside of the 120 incorporated municipalities) within CCDPH is estimated to be about 104,874, or 4.6 percent of the jurisdiction’s total. 

```{r}
#| label: fig-sccmunis
#| fig-cap: Municipalities that are either partially or wholly located in Cook County
#| warning: false
#| eval: true

con_inter_census <- dbConnect(
  odbc::odbc(),
  Driver   = "SQL Server",
  Server   = key_get("ccdph_sql_server"),
  Database = "inter-census"
)

pop_munis_cook_county <- dbReadTable(conn = con_inter_census,
                               Id(schema = "ref", 
                                  table = "decennial-2020-total-by-muni"))

munis_cook_county <- dbReadTable(conn = con_inter_spatial,
                               Id(schema = "ref", 
                                  table = "munis_cook_county"))

munis_cook_county_frm <- munis_cook_county %>%
  st_drop_geometry() %>%
  left_join(pop_munis_cook_county %>% select(geoid_place, pop_total, pop_in_cook), by="geoid_place") %>%
  rename(name = name_place,
         geoid = geoid_place,
         partial = partial_ccdph,
         sqmi = area_sqmi) %>%
  select(-c(area_land_sqmi:area_water_sqmi, geom, partial, exclude))

munis_cook_county_frm %>%
  arrange(name) %>%
    datatable(class = 'cell-border stripe',
            rownames = FALSE,
            options = list(pageLength = 5),
            filter = "top") %>%
  formatRound('sqmi',1) %>%
  formatRound(c(8:9),0, interval = 3, mark =  ",") %>%
  formatStyle(columns = c(1:9),
              fontSize = '75%')

```

# Accessing geographic data

For convenience, the Epidemiology Unit's base or reference geographic layers are stored in three locations using the same naming convention.

1. [CCDPH SQL Server](storage.qmd#sec-ccdph_sql_server_storage) *inter-spatial* database: Geographic data stored on the SQL Server are considered to be the definitive source of spatial data. They are continually updated and accessed by the Epidemiology and Communicable Disease units. SQL Server, due to its robust querying functionality, also offers the most flexible and efficient means to manage, write, store, and retrieve geographic data. The inter-spatial data can also be easily imported into SAS, R, ArcGIS Pro, and Microsoft applications such as Excel and Access.

::: {.callout-note collapse="false"}
## Code for reading and writing geographic layers from/to CCDPH SQL Server via R

To import the read and write spatial layer to database functions into your R script, copy, paste, then run the following code in R. 

```{verbatim}
source("https://raw.githubusercontent.com/Cook-County-Department-of-Public-Health/ccdph-functions/master/spatial-database-functions.R")
```

To read spatial data into R use the *fx_read_spatial_layer_fr_database* function. Parameters for this function include: *schema_name* or name of the schema that the spatial database table is assigned to (defaults to "ref"); *db_table_name* or name of spatial table in inter-spatial database (refer to @fig-spatialtables above); and *crs_id* or the desired coordinate reference system  (defaults to 3435 but set to 4326 when mapping in leaflet). The example code below imports the clipped municipalities in Cook County data layer into R as a *munis_cook_county_clipped* simple feature dataframe, reprojected into geographic coordinate system 4326.

```{verbatim}

munis_cook_county_clipped <- fx_read_spatial_layer_fr_database(
  db_table_name = "munis_cook_county_clipped", 
  crs_id = 4326
  )

```

To write spatial data into R use the *fx_read_spatial_layer_to_database* function, which has the following parameters: *sf_layer_df* or the simple feature data frame you wish to write to the database; *schema_name* or name of the schema that the spatial database table will be assigned to (defaults to "ref"); *db_table_name* or name of spatial table that will be written to the database (WARNING: this will overwrite any existing table with the same name); and *crs_id* or the desired coordinate reference system for the spatial table to be written to the database (defaults to 3435). The example code below writes the counties in Illinois simple feature dataframe in R to a spatial table in the spatial database with name *counties_illinois* assigned to schema, *ref*.

```{verbatim}

fx_write_spatial_layer_to_database(
  sf_layer_name = counties_il_sf,
  schema_name = "ref",
  db_table_name = "counties_illinois",
  crs_id = 3435
)

```

Be sure to update the *00_table_dictionary* reference table in R when writing new spatial layers to the database.

:::

2. [ccdph-shapefiles GitHub repository](https://github.com/Cook-County-Department-of-Public-Health/ccdph-shapefiles): The geographic layers stored on SQL Server are also exported to the *ccdph-shapefiles* GitHub repository as zip files. These zip files are used for importing spatial data into ArcGIS Online and also for sharing with other internal and external entities.

3. [ArcGIS Online in the CCDPH Epidemiology (Internal) group content area](https://cookcountyil.maps.arcgis.com/home/group.html?id=762568106bd341f0a14c942dcca890e7#overview): Feature layers stored on Cook County's ArcGIS Online can be easily imported into ArcGIS Pro for mapping. The same feature layers are used to create map templates. At this time, Cook County's ArcGIS Online is not able to read directly from the CCDPH SQL Server, so data stored here were imported from zipped shapefiles stored on the *ccdph-shapefiles* GitHub repository. 
