---
title: "RACE AND ETHNICITY CATEGORIES EMPLOYED BY CCDPH'S SURVEILLANCE PROGRAMS"
subtitle: "<i><em>Insights</em> concerning CCDPH's current and proposed collection and analysis of race and ethnicity data</i>"
date: "`r format(Sys.time(), '%d %B %Y')`"
  
format:
  html:
    theme: [journal, "../reports.scss"]
    favicon: "https://cookcountypublichealth.org/wp-content/uploads/2018/12/cropped-favicon-32x32.png"
    code-tools: false  
    code-fold: false
    echo: false
    title-block-banner: true
    toc: true
    toc-depth: 4
    toc-location: left
    include-in-header: "../header.html"
    reference-location: document
    fig-cap-location: top
    fontsize: 0.9em
    embed-resources: true
bibliography: "../references.bib"
editor_options: 
  chunk_output_type: console
---

```{r}
#| label: activate R packages
#| echo: false
#| warning: false

library(keyring)
library(dplyr)
library(data.table)
library(DT)
library(tidyverse)
library(openxlsx)
library(censusapi)
library(tigris)

setwd(paste0("E:/OneDrive - Cook County Health/git_repos/justenvirons/ccdph-documents/documents/reports/race-ethnic-data-management"))

# acensusapikey <- key_get("censusapi")
```

# Introduction

This document provides a brief overview of the race and ethnic categories utilized by the Cook County Department of Public Health (CCDPH) Epidemiology and Communicable Disease Units as part of their data management activities. It provides detailed information concerning the particular race and ethnic categories available in specific datasets, how these data are standardized to calculate rates, where these data are disseminated, and considerations for CCDPH as it plans to transition to new/proposed categorization schema that take into account Middle Eastern, North African (MENA), and/or South Asian/Indian populations.

The CCDPH uses race and ethnic categories to stratify public health data both in the collection of primary data (e.g., surveys) as well as in the evaluation of data made available via secondary sources (e.g., Illinois Department of Public Health). The following section summarizes specific datasets used by CCDPH in its surveillance activities and the race and ethnic categories available in each dataset. Refer to the searchable table in the [Comprehensive Race and Ethnicity Table] section of this document.

The datasets summarized below are organized into two broad sections: *primary* and *secondary*. It should be noted that *CCDPH has more control in defining and framing race and ethnic categories in its **primary data** collection efforts and less agency in altering categories in the **secondary data** which are curated by external entities* (e.g., IDPH)*.* Further, transitioning to a more nuanced accounting of race and ethnicity (i.e., including, but not limited to MENA populations) may not be easily translated into respective rate calculations in the short-term given that such rates rely on population counts and estimates provided by the US Census Bureau. These and other considerations, unintended consequences, and recommendations are briefly addressed in the [Discussion] section of this document.

# Primary data

CCDPH's Epidemiology Unit engages in two primary data collection efforts: (1) the Cook County Health Survey (CCHS), administered annually; and (2) the Youth Risk Behavior Survey (YRBS), administered every two years. The CCDPH has discretion over final design of these surveys, including questions concerning race and ethnicity.

### Cook County Health Survey (CCHS)

The [Cook County Health Survey (CCHS)](https://cookcountypublichealth.org/health-survey/) is a cross-sectional survey with a sample of approximately 32,000 randomly selected adults aged 18 and older suburban Cook County residents. The paper and web-based versions of the survey are available in English and Spanish. All data collected are self-reported and used to help the CCDPH assess the health-related needs of the county population and to guide the Department in planning programs to improve the health and well being of county residents. The design of the survey including question phrasing and structure closely follows [CDC's national Behavioral Risk Factor Surveillance System](https://www.cdc.gov/brfss/data_documentation/index.htm) (BRFSS) and the [Illinois County Behavioral Risk Factor Surveys](http://www.idph.state.il.us/brfss/countydata.asp?selTopicCounty=R6Overview&areaCounty=Cook_90.1&show=freq&yrCounty=6&form=county&yr=&area=&selTopic=) (ICBRFS) programs.

The CCDPH has traditionally sought to align its survey with validated questions asked in the BRFSS, ICBRFS, as well as the similarly structured [Healthy Chicago Survey](https://www.chicago.gov/city/en/depts/cdph/supp_info/healthy-communities/healthy-chicago-survey.html) (HCS) via a broader collaborative effort with the Chicago Department of Public Health (CDPH). The 2022 version of the CCHS included a series of questions concerning the race and ethnicity of respondents (see data table below). The latest survey was administered 2022. In 2023, the CCDPH Epidemiology Unit will administer a revised health survey, the questions for which have been recently finalized via a weeks-long collaborative process. The CCHS is slated to be carried out annually in upcoming years.

```{r}
#| label: tbl-cchscategories
#| tbl-cap: 2022 CCHS Detailed Race and Ethnic Data
#| warning: false

cchs_cats <- read.xlsx(xlsxFile = "data/race-ethnic-data-management-input-table-ccdph.xlsx", sheet="Primary_CCHS")

cchs_cats_table <- cchs_cats %>% 
  rename("Categories" = 3) %>%
  slice_head(n=19) %>%
  group_by(Question, Order) %>%
  summarise(Categories = toString(`Categories`)) %>%
  arrange(Order) %>%
  select(-c(Order)) %>%
  mutate(Type = "Primary Data",
         Source = "Cook County Health Survey",
         Owner = "CCDPH",
         Program = "Epidemiology") %>%
  select(Type, Source, Owner, Program, Question, Categories)

cchs_cats_table %>% 
  select(Question, Categories) %>%
    datatable(class = 'cell-border stripe',
            rownames = FALSE,
            options = list(pageLength = 10,
                           dom="t")) %>%
  formatStyle(columns = c(1, 2),
              fontSize = '75%')

```

### Youth Risk Behavior Survey (YRBS)

The CCDPH collects race and ethnicity data for younger populations via its [Youth Risk Behavior Survey (YRBS)](https://cookcountypublichealth.org/epidemiology-data-reports/yrbs-youth-risk-behavioral-survey/). Administered in 2010 and 2020 with plans to administer every two years going forward, the survey assesses health risk behaviors among students (grade 9-12) in participating high schools throughout suburban Cook County. The survey is used to monitor and evaluate a variety of health-related behaviors and conditions among adolescents, including physical activity, sexual behavior, nutrition, alcohol and drug use, and mental health. Survey results can be used to inform the development of health policies and programs that address the needs of adolescents. The 2020 and 2022 versions of the YRBS included a series of questions concerning the race, ethnicity and ancestry of respondents (see below data table).

```{r}
#| label: tbl-yrbscategories
#| tbl-cap: Race and Ethnic Data Collected in 2020 and 2022 YRBS
#| warning: false

yrbs_cats <- read.xlsx(xlsxFile = "data/race-ethnic-data-management-input-table-ccdph.xlsx", sheet="Primary_YRBS")

yrbs_cats_table <- yrbs_cats %>% 
  rename("Categories" = 2) %>%
  slice_head(n=7) %>%
    group_by(Question) %>%
  summarise(Categories = toString(`Categories`)) %>%
  mutate(Type = "Primary Data",
         Source = "Cook County Health Survey",
         Owner = "CCDPH",
         Program = "Epidemiology") %>%
  select(Type, Source, Owner, Program, Question, Categories)

 yrbs_cats_table %>%
   select(Question, Categories) %>%
   datatable(class = 'cell-border stripe',
            rownames = FALSE,
            options = list(pageLength = 10,
                           dom="t")) %>%
  formatStyle(columns = c(1,2),
              fontSize = '75%')
```

# Secondary data

## Vital statistics

The CCDPH receives vital statistics (i.e., births, deaths) data on a regular basis from the Illinois Department of Public Health (IDPH) [Vital Statistics Office](https://dph.illinois.gov/data-statistics/vital-statistics.html). The CCDPH Epidemiology Unit processes, analyzes, and disseminates these data as rates on the [Cook County Health Atlas](https://cookcountyhealthatlas.org/), custom data requests, and through other data briefs and reports.

### Mortality

The IDPH makes available [mortality data](https://dph.illinois.gov/data-statistics/vital-statistics/death-statistics.html) with extensive information on the decedent's cause of death and demographic characteristics. See the below table for a list of race and ethnic categories used in this dataset. The Epidemiology Unit requests the latest data from IDPH on an annual basis.

```{r}
#| label: tbl-deathscategories
#| tbl-cap: Race and Ethnic Data Collected in 2020 IDPH Death Statistics
#| warning: false

deaths_cats <- read.xlsx("data/race-ethnic-data-management-input-table-ccdph.xlsx", sheet="Secondary_Mortality")

deaths_cats_table <- deaths_cats %>% 
  rename("Question" = 1,
         "Categories" = 2) %>%
  slice_head(n=21) %>%
  group_by(Question) %>%
  summarise(Categories = toString(`Categories`)) %>%
  mutate(Type = "Secondary Data",
         Source = "Vital Records: Mortality",
         Owner = "IDPH",
         Program = "Epidemiology") %>%
  select(Type, Source, Owner, Program, Question, Categories)

deaths_cats_table %>%
  select(Question, Categories) %>%
  datatable(class = 'cell-border stripe',
            rownames = FALSE,
            options = list(pageLength = 10,
                           dom="t")) %>%
  formatStyle(columns = c(1,2),
              fontSize = '75%')

```

<br />

### Natality

Similar to mortality, the IDPH shares [natality data](https://dph.illinois.gov/data-statistics/vital-statistics/birth-statistics.html) with extensive information on the race and ethnicity of the mother and father. See the below table for a list of race and ethnic categories used in this dataset. The Epidemiology Unit requests the latest data from IDPH on an annual basis.

```{r}
#| label: tbl-birthscategories
#| tbl-cap: Race and Ethnic Data Collected in 2020 IDPH Birth Statistics
#| warning: false

births_cats <- read.xlsx("data/race-ethnic-data-management-input-table-ccdph.xlsx", sheet="Secondary_Natality")

births_cats_table <- births_cats %>% 
  rename("Question" = 1,
         "Categories" = 2) %>%
  slice_head(n=42) %>%
  group_by(Question) %>%
  summarise(Categories = toString(`Categories`)) %>%
  mutate(Type = "Secondary Data",
         Source = "Vital Records: Natality",
         Owner = "IDPH",
         Program = "Epidemiology") %>%
  select(Type, Source, Owner, Program, Question, Categories)

births_cats_table %>%
    select(Question, Categories) %>%
    datatable(class = 'cell-border stripe',
            rownames = FALSE,
            options = list(pageLength = 10,
                           dom="t")) %>%
  formatStyle(columns = c(1,2),
              fontSize = '75%')

```

## Hospitalization Discharge

Hospital discharge inpatient and outpatient data are administrative data collected when a patient is discharged from a hospital. Inpatient data is collected on patients admitted to a hospital for an overnight stay, while outpatient data is collected on patients who receive services from a hospital but do not stay overnight. The CCDPH receives both [inpatient and outpatient data](https://dph.illinois.gov/topics-services/prevention-wellness/patient-safety-quality/discharge-data.html) (which includes patient demographic information, diagnosis codes, procedures codes, and charges) from the IDPH on a regular basis. The Epidemiology Unit cleans, processes and creates population-standardized rates of hospital discharges for select diagnoses by zip code. These rates are then made available on the [Cook County Health Atlas](https://cookcountyhealthatlas.org/) among other data briefs and reports. Refer to the below table for a list of race and ethnic categories used in this dataset.

```{r}
#| label: tbl-hospitalcategories
#| tbl-cap: Race and Ethnic Data Collected in 2020 IDPH Death Statistics
#| warning: false

discharge_cats <- read.xlsx("data/race-ethnic-data-management-input-table-ccdph.xlsx", sheet="Secondary_HospitalDischarge")

discharge_cats_table <- discharge_cats %>% 
  rename("Question" = 1,
         "Categories" = 2) %>%
  slice_head(n=11) %>%
  group_by(Question) %>%
    summarise(Categories = toString(`Categories`)) %>%
  mutate(Type = "Secondary Data",
         Source = "Hospital Discharge",
         Owner = "IDPH",
         Program = "Epidemiology") %>%
  select(Type, Source, Owner, Program, Question, Categories)

discharge_cats_table %>%
  select(Question, Categories) %>%
  datatable(class = 'cell-border stripe',
            rownames = FALSE,
            options = list(pageLength = 10,
                           dom="t")) %>%
  formatStyle(columns = c(1,2),
              fontSize = '75%')

```

## Emergency Medical Services (EMS)

Emergency Medical Services (EMS) data are collected from the activities of emergency medical services such as ambulances, hospitals, and emergency rooms. These data include information such as patient demographics (see race and ethnic categories below), diagnoses, treatment protocols, and outcome data. CCDPH accesses near real time [biospatial EMS data](https://www.biospatial.io/about/) via an arrangement with IDPH. The data are reported at multiple geographic levels including zip code and are used by CCDPH's Epidemiology Unit to identify trends and spatial patterns in emergency care. The EMS data made available via biospatial are stratified by race and/or ethnic group for larger areas (CCDPH jurisdiction, health district) but not smaller geographies due to patient privacy concerns and data suppression guidelines.

```{r}
#| label: tbl-emscategories
#| tbl-cap: Race and Ethnic Categories Collected in EMS Data 
#| warning: false

ems_cats <- read.xlsx("data/race-ethnic-data-management-input-table-ccdph.xlsx", sheet="Secondary_EMS")

ems_cats_table <- ems_cats %>% 
  rename("Question" = 1,
         "Categories" = 2) %>%
  slice_head(n=7) %>%
  group_by(Question) %>%
  summarise(Categories = toString(`Categories`)) %>%
  mutate(Type = "Secondary Data",
         Source = "EMS/Biospatial",
         Owner = "IDPH",
         Program = "Epidemiology") %>%
  select(Type, Source, Owner, Program, Question, Categories)
  
ems_cats_table %>%
  select(Question, Categories) %>%
    datatable(class = 'cell-border stripe',
            rownames = FALSE,
            options = list(pageLength = 10,
                           dom="t")) %>%
  formatStyle(columns = c(1,2),
              fontSize = '75%')

```

## Behavioral Risk Surveys

The [Illinois County Behavioral Risk Factor Survey (ICBRFS)](http://www.idph.state.il.us/brfss) and [Centers for Disease Control (CDC) BRFSS (Behavioral Risk Factor Surveillance System)](https://www.cdc.gov/brfss/index.html) are annual state-based telephone surveys of the civilian, non-institutionalized population. The surveys are used to collect data about health-related behaviors, chronic health conditions, and use of preventive services. Data are collected through interviews with a representative sample of adults (aged 18 years and older), which are then aggregated and reported for various geographies. Responses to the ICBRFS are reported as aggregated (both weighted and unweighted) counts and prevalence rates for suburban Cook County as a whole. The national BRFSS data are reported by county. The CDC also uses BRFSS data to estimate prevalence by health indicator for small geographic areas (i.e., census tracts, zip code tabulation areas, municipalities) throughout the country as part of its [PLACES program](https://www.cdc.gov/places/).

The CCDPH utilizes ICBRFS, BRFSS and PLACES data to identify trends and patterns in self-reported health conditions as part of its Community Health Status Assessment (CHSA). CCDPH also publishes PLACES data on its [Cook County Health Atlas](https://cookcountyhealthatlas.org/). Note that while these data programs collect demographic information about survey respondents (see below table), the information is not used to stratify prevalence rates by race and/or ethnic group.

```{r}
#| label: tbl-brfsscategories
#| tbl-cap: Race and Ethnic Data Collected in ICBRFS and BRFSS Surveys
#| warning: false

brfss_cats <- read.xlsx("data/race-ethnic-data-management-input-table-ccdph.xlsx", sheet="Secondary_BRFSS")

brfss_cats_table <- brfss_cats %>% 
  rename("Question" = 2,
         "Categories" = 3) %>%
  slice_head(n=43) %>%
  group_by(Question, Order) %>%
  arrange(Order) %>%
  select(-c(Order)) %>%
  summarise(Categories = toString(`Categories`)) %>%
  mutate(Type = "Secondary Data",
         Source = "ICBRFS/BRFSS",
         Owner = "IDPH/CDC",
         Program = "Epidemiology") %>%
  select(Type, Source, Owner, Program, Question, Categories)

brfss_cats_table %>%
  select(Question, Categories) %>%
    datatable(class = 'cell-border stripe',
            rownames = FALSE,
            options = list(pageLength = 10,
                           dom="t")) %>%
  formatStyle(columns = c(1,2),
              fontSize = '75%')

```

## Illinois' National Electronic Disease Surveillance System (I-NEDSS)

Data on reportable infectious diseases collected by local health departments are stored in a web-based surveillance application, [I-NEDSS](https://dph.illinois.gov/topics-services/diseases-and-conditions/infectious-diseases/infectious-disease-reporting.html). Race and ethnic information is populated by either the provider or laboratory reporting the case, or by CCDPH staff upon case investigation. Data available in I-NEDSS may be exported by authorized users for a given health jurisdiction at any time, with the exception of HIV and syphilis case reports. Case reports for these diseases are stored in separate systems and provided from IDPH to local health jurisdictions on an annual basis. Refer to the table below for race and ethnic categories available for reportable infectious diseases.

```{r}
#| label: tbl-inedsscategories
#| tbl-cap: Race and Ethnic Data Collected in I-NEDSS System
#| warning: false

inedss_cats <- read.xlsx("data/race-ethnic-data-management-input-table-ccdph.xlsx", sheet="Secondary_INEDSS")

inedss_cats_table <- inedss_cats %>% 
  rename("Question" = 1,
         "Categories" = 2) %>%
  slice_head(n=9) %>%
  group_by(Question) %>%
  summarise(Categories = toString(`Categories`)) %>%
  mutate(Type = "Secondary Data",
         Source = "INEDSS",
         Owner = "IDPH",
         Program = "Communicable Disease") %>%
  select(Type, Source, Owner, Program, Question, Categories)

inedss_cats_table %>%
  select(Question, Categories) %>%
    datatable(class = 'cell-border stripe',
            rownames = FALSE,
            options = list(pageLength = 10,
                           dom="t")) %>%
  formatStyle(columns = c(1,2),
              fontSize = '75%')

```

## Illinois Comprehensive Automated Immunization Registry Exchange (I-CARE)

[I-CARE](https://dph.illinois.gov/topics-services/prevention-wellness/immunization/icare.html) is a web-based immunization record-sharing application developed by IDPH to enable immunization record sharing between healthcare providers, statewide. Local health departments may access I-CARE to look up immunization records for cases under investigation for reportable diseases. For some diseases, such as COVID-19, local health departments may receive data reports from I-CARE to assist in assessing immunization coverage for their jurisdiction. Race and ethnic information is populated by providers reporting immunizations to I-CARE. Refer to the table below for race and ethnic categories available for immunization reporting.

```{r}
#| label: tbl-icarecategories
#| tbl-cap: Race and Ethnic Data Collected in I-CARE System
#| warning: false

icare_cats <- read.xlsx("data/race-ethnic-data-management-input-table-ccdph.xlsx", sheet="Secondary_ICARE")

icare_cats_table <- icare_cats %>% 
  rename("Question" = 1,
         "Categories" = 2) %>%
  slice_head(n=12) %>%
  group_by(Question) %>%
  summarise(Categories = toString(`Categories`)) %>%
  mutate(Type = "Secondary Data",
         Source = "I-CARE",
         Owner = "IDPH",
         Program = "Communicable Disease") %>%
  select(Type, Source, Owner, Program, Question, Categories)

icare_cats_table %>%
  select(Question, Categories) %>%
    datatable(class = 'cell-border stripe',
            rownames = FALSE,
            options = list(pageLength = 10,
                           dom="t")) %>%
  formatStyle(columns = c(1,2),
              fontSize = '75%')

```

## Biosense Platform, National Syndromic Surveillance Program (NSSP)

Acute care hospitals in Illinois are required report selected information on emergency room encounters and inpatient admissions to IDPH. These data are typically referred to as syndromic data. IDPH submits syndromic data to [CDC's Biosense Platform](https://www.cdc.gov/nssp/overview.html), part of the National Syndromic Surveillance Program. Authorized users from local health departments may access the Biosense Platform to conduct syndromic surveillance activities, such as monitoring trends in opioid overdoses, extreme weather events, or emerging public health threats. Race and ethnic information in the platform is populated from patients' electronic health records. Refer to the table below for race and ethnic categories available for syndromic records.

```{r}
#| label: tbl-nsspcategories
#| tbl-cap: Race and Ethnic Data Collected in NSSP Program
#| warning: false

nssp_cats <- read.xlsx("data/race-ethnic-data-management-input-table-ccdph.xlsx", sheet="Secondary_NSSP")

nssp_cats_table <- nssp_cats %>% 
  rename("Question" = 1,
         "Categories" = 2) %>%
  slice_head(n=17) %>%
  group_by(Question) %>%
  summarise(Categories = toString(`Categories`)) %>%
  mutate(Type = "Secondary Data",
         Source = "NSSP",
         Owner = "IDPH",
         Program = "Communicable Disease") %>%
  select(Type, Source, Owner, Program, Question, Categories)

nssp_cats_table %>%
  select(Question, Categories) %>%
    datatable(class = 'cell-border stripe',
            rownames = FALSE,
            options = list(pageLength = 10,
                           dom="t")) %>%
  formatStyle(columns = c(1,2),
              fontSize = '75%')

```

## US Census Bureau Population Counts and Estimates

The CCDPH Epidemiology and Communicable Disease Units construct standardized rates using population counts made available from the US Census Bureau; specifically its [decennial census counts](https://www.census.gov/programs-surveys/decennial-census.html) and [American Community Survey](https://www.census.gov/programs-surveys/acshttps://www.census.gov/programs-surveys/acs) 1-, 3-, and 5-year estimates. These datasets provide detailed population counts by race, ethnicity as well as other groupings (e.g., age, sex) commonly used to stratify rates. Race and ethnicity categories employed in the US Census have been modified over the decades to reflect the evolving demography of the country, although full understanding of the meaning of these constructs need ongoing refinement and study (hence the proposed MENA category).

In 1997, the Office of Management and Budget (OMB) issued the latest [Revisions to the Standards for the Classification of Federal Data on Race and Ethnicity](https://obamawhitehouse.archives.gov/omb/fedreg_1997standards), which are used for federal data collection purposes in the US census, other national household surveys, and in medical and clinical research. The five current categories for race are defined as: American Indian or Alaska Native, Asian, Black or African American, Native Hawaiian and Other Pacific Islander, and White. A multirace category was included for the first time in the 2000 census and the Epidemiology and Communicable Disease Units have agreed to include a "multiracial" category in all relevant reporting. (Between 2000 and 2020, the multiracial resident category was the fastest growing population subgroup in CCDPH's jurisdiction. Refer to the Cook County Health Atlas for [current demographic characteristics](https://cookcountyhealthatlas.org/indicators/PCT?topic=population-percentages-by-demographic-group). There are also two categories for ethnicity: "Hispanic or Latino" and "Not Hispanic or Latino" and the question of ethnicity is posed before the race categories. The 2020 census asked about family background for all respondents and continues to collect information on nation or US territory of origin for Hispanics/Latinos, Other Pacific Islanders and Asians, as well as tribal affiliation for American Indians/Alaska Natives. These standard categories are an essential component of minority and health disparities research and are considered the standard nomenclature in the public health field.

# Discussion

CCDPH and other institutions are presently very limited in how they can report out health conditions for its MENA population given that people of Middle East and North African descent have been largely forced to choose either Asian or White in conventional public health and population data collection efforts. While it may be a good idea for CCH and CCDPH to begin asking/collecting such data in their own surveys including the latest Cook County Health Survey, doing so would also require changes in institutional practices. Considerations, unintended consequences, and proposals concerning such a transition are briefly summarized below.

## Considerations

-   *Calculating rates*: CCDPH regularly calculates and reports disease, health outcomes, behaviors, and conditions as both counts (i.e., frequencies over a given time period) and rates or proportions (i.e., standardized frequencies over a given time period). These counts and rates take the form of incidence and prevalence, which measure the number of new cases and the total number of existing cases of a condition, respectively.

    Rates are especially useful when comparing conditions across locations and population subgroups, including by race and ethnicity. Calculating rates requires dividing the frequency of a given condition by the size of the reference population. The condition-specific frequencies (i.e., the numerator) are collected via the various sources summarized above whereas the reference resident population totals (i.e., the denominator) is made available from the US Census Bureau as part of its decennial [census of Population and Housing](https://www.census.gov/programs-surveys/decennial-census.html) and/or its 1-, 3- and 5-year [American Community Survey](https://www.census.gov/programs-surveys/acs) (ACS). For example, calculating crude rates for CCDPH's Black, Not Latino/a resident population requires the number of subgroup-specific events (e.g., cancer-related death or hospitalization) divided by the sum of the total Black, Not Latino/a population residing within CCDPH's jurisdiction. Such rates are useful for estimating actual risk of hospitalization and mortality and documenting disparities across race and ethnic populations with similar age distributions.

    Calculating MENA-specific rates becomes problematic because the census does not presently collect and disseminate population counts and/or estimates for that population subgroup in its standard race and ethnicity tables. That said, even if CCDPH decides to revise its demographic questions (as part of its CCHS and YRBS, say) to include a MENA race, ethnic category, it would only be able to report counts of conditions for that group, not rates given the lack of MENA-specific population counts/estimates from the US Census.

-   *Timing*: The Biden administration recently proposed adding MENA as a category to the 2030 census. CCDPH would likely need to continue using conventional race and ethnic categories in their epidemiological rate calculations until MENA population counts and estimates are made available.

## Unintended Consequences

-   *Politics*: Whether or not to include the MENA category has been a contentious issue for the US Census Bureau over the past several years. In 2018, under the Trump administration, the bureau declined to add the category. Further, there may be concerns among stakeholders about revising the current race and/or ethnic population categories given that the new structure may reduce populations, shares, and perhaps political/power influence of conventionally-defined groups.

## Proposals, Next Steps

-   Consistent reporting out of epidemiological counts and rates for existing multiracial (i.e., two or more races) category in the standard race and ethnicity tables. This category may help account for individuals who do not fully identify with any single racial group;

-   Work with CCH and other CCDPH units to establish a plan for transitioning to proposed MENA category, including the identification of additional considerations, potential unintended consequences, and proposals; and

-   Submit a formal comment in support of [Initial Proposals For Updating OMB's Race and Ethnicity Statistical Standards](https://www.federalregister.gov/documents/2023/01/27/2023-01635/initial-proposals-for-updating-ombs-race-and-ethnicity-statistical-standards) by the 4/12/2023 deadline.

# Comprehensive Race and Ethnicity Table

Below is a searchable list of race and ethnic categories used by the CCDPH Epidemiology and Communicable Disease Units by type (i.e., primary, secondary), source, owner, and program. Data for table can be downloaded as [Excel spreadsheet](data/race-ethnic-data-management-output-table-ccdph.xlsx) or delimited [CSV file](data/race-ethnic-data-management-output-table-ccdph.csv).

```{r}
#| label: tbl-compcategories
#| tbl-cap: Race and Ethnic Categories by Source, Owner and Program
#| warning: false

alist_raceethnic_tables <- mget(ls(pattern = "_table"))
alist_raceethnic_tables_all <- rbindlist(alist_raceethnic_tables, fill=TRUE)

alist_raceethnic_tables_all %>%
  arrange(Type,Source,Owner,Program) %>%
    datatable(class = 'cell-border stripe',
            rownames = FALSE,
            options = list(pageLength = 30)) %>%
  formatStyle(columns = c(1:6),
              fontSize = '75%')

write_csv(alist_raceethnic_tables_all, "data/race-ethnic-data-management-output-table-ccdph.csv")
write.xlsx(alist_raceethnic_tables_all,"data/race-ethnic-data-management-output-table-ccdph.xlsx")



```
